---
title: "class06"
author: 'Delaney (PID: A15567985)'
date: "2/3/2022"
output:
  pdf_document: default
  html_document: default
---

Today we will explore R functions. 

We will start with calculating a grade for these example students. 
```{r}
# Example input vectors to start with
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
```

We could use the mean() function to calculate an average. 

```{r}
mean(student1)
mean(student2, na.rm= TRUE)
```

How does the is.na() function work? Let's try it out on student2

```{r}
student2
is.na(student2)
```

We can use this result to get our NA values (i.e. the TRUE positions)

```{r}
student2[is.na(student2)]
```

We can set the NA value to 0 and find the average grade. 

```{r}
student2[is.na(student2)] <- 0
student2
mean(student2)
```

```{r}
student3[is.na(student3)] <- 0
student3
mean(student3)
```

Now we want to find the lowest score... 
```{r}
student1
min(student1)
which.min(student1)
```

The which.min() function tell us where the smallest valie is. 
```{r}
student1[ -which.min(student1)]
```

Average grade minus lowest score. 

```{r}
x <- student2
x <- student2[ -which.min(x)]
mean(x)
```

We are close to our working code snippet that will be the body of our first function. 

```{r}
# First set NA values to zero
x[is.na(x)] <- 0 
# Remove lowest score and calculate average 
mean(x[ -which.min(x)])
```

Now we can turn this into our first function. We can call this function 'grade()'.

All R functions have 3 things
- a name (grade)
- input arguments (student scores)
- body (does the work!)

```{r}
grade <- function(x) {
  # First set NA values to zero
  x[is.na(x)] <- 0 
  # Remove lowest score and calculate average 
  mean(x[ -which.min(x)]) 
}
```

Let's try it out. 
```{r}
grade(student2)
```

> Q1. Write a function grade() to determine an overall grade from a vector of student homework
assignment scores dropping the lowest single score. If a student misses a homework (i.e. has an
NA value) this can be used as a score to be potentially dropped. Your final function should be
adquately explained with code comments and be able to work on an example class gradebook
such as this one in CSV format: “https://tinyurl.com/gradeinput”

```{r}
gradebook <- read.csv("https://tinyurl.com/gradeinput", row.names=1)
head(gradebook)
```

We can use the 'apply()' function to grade the whole class. 

```{r}
scores <-apply(gradebook, 1, grade)
scores
```

>Q2. Using your grade() function and the supplied gradebook, Who is the top scoring student
overall in the gradebook?

```{r}
which.max(scores)
```

>Q3. From your analysis of the gradebook, which homework was toughest on students (i.e. obtained the lowest scores overall?

```{r}
hw.mean <- apply(gradebook, 2, median, na.rm= TRUE)
which.min(hw.mean)
```

```{r}
boxplot(gradebook)
```


>Q4. From your analysis of the gradebook, which homework was most predictive of overall score (i.e. highest correlation with average grade score)?

```{r}
cor(scores, gradebook$hw1)
```

Now apply over all columns (i.e. homeworks)

```{r}
apply(gradebook, 2, cor, scores)
```

Change all NA to 0 
```{r}
mask <- gradebook
mask[ is.na(mask)] <- 0
```

Use mask instead of gradeboo
```{r}
apply(mask, 2, cor, scores)
```


>Q5. Make sure you save your Rmarkdown document and can click the “Knit” button to generate
a PDF foramt report without errors. Finally, submit your PDF to gradescope.