---
title: 'Class 10: Genome Informatics'
author: 'Delaney (PID: A15567985)'
date: "2/17/2022"
output:
  pdf_document: default
  html_document: default
---

# Examine 1000 Genome Data 

>Q5: What proportion of the Mexican Ancestry in Los Angeles sample population (MXL) are homozygous for the asthma associated SNP (G|G)?

```{r}
# Read genotype file from Ensemble
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
table(mxl$Genotype..forward.strand.)/nrow(mxl)
```

What about a different population? Here we take the British in England and Scotland (GBR)
```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
table(gbr$Genotype..forward.strand.)/nrow(gbr)
```


## Expression by Genotype Analysis

I want to read my RNA-Seq expression results into R. This file is not a CSV but rather has fields seperated by space. 

```{r}
x <- read.table("geneexpression.txt")
head(x)
```

First try at this question. Is the mean expression different based on genotype?
```{r}
x$geno == "G/G"
```

```{r}
summary(x[x$geno == "G/G", 3])
```

Now we will look at other genotypes.
```{r}
summary(x[x$geno == "A/A", 3])
```

```{r}
summary(x[x$geno == "A/G", 3])
```

## Make a summary overview figure 

Make a boxplot figure...
```{r}
library(ggplot2)
ggplot(x) + aes(geno, exp, fill=geno) + geom_boxplot(notch= TRUE)
```

