---
title: "Machine Learning 1"
author: 'Delaney (PID: A15567985)'
date: "2/8/2022"
output:
  pdf_document: default
  html_document: default
---

# First up kmeans()

Demo od using kmeans() function in base R. First make up some data with a known structure

```{r}
tmp <- c( rnorm(30, -3), rnorm(30, 3) )
x <- cbind(tmp, y=rev(tmp))
plot(x)
```

Now we have some made up data in 'x' lets see how kmeans works with this data

```{r}
k <- kmeans(x, centers=2, nstart=20)
k
```

>Q. How many points are in each cluster?

```{r}
k$size
```

>Q. How do we go to the cluster membership/assignment?

```{r}
k$cluster
```

>Q. What about cluster centers?

```{r}
k$centers
```

Now we've got to the main results let's use them to plot our data with kmeans result. 

```{r}
plot(x, col=k$cluster)
points(k$centers, col="blue", pch=15)
```

## Now for Hierachical Clustering

We will cluster the same data 'x' with the hclust(). In this case 'hclust()' requires a distance matrix as input. 

```{r}
hc <- hclust(dist(x))
hc
```

Let's plot our hclust result

```{r}
plot(hc)
```

To get our cluster membership vector we need to "cut" the tree with the 'cutree()'. 

```{r}
grps <- cutree(hc, h=8)
grps
```

Now plot our data with the hclust() results. 

```{r}
plot(x, col=grps)
```

# Principal Component Analysis (PCA)

## PCA of UK food data

Read data from website and try a few visualizations. 

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url, row.names=1)
x
```
## Data Import

>Q1. How many rows and columns are in your new data frame named 'x'? Whar R functions could you use to answers this question?

```{r}
dim(x)
```

There are 17 rows and 4 columns for the new data frame named 'x'. The R functions that can be used to answer this are 'dim(x)' for both or 'ncol()' and 'nrow()' seperately.

## Checking your data 

>Q2. Which approach to solving the 'rownames problem' mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

I prefer using the argument setting 'row.names()' to set the correct row-names because it only returned the data frame with the row names changed. 'Rownames()' is a non-generic function, while 'row.names()' is a generic function and is specific for data frames. 

```{r}
cols <- rainbow(nrow(x))
barplot( as.matrix(x), col=cols )
```

## Spotting major differences and trends 

```{r}
barplot( as.matrix(x), col=cols, beside=TRUE )
```

>Q3. Changing what optional argument in the above barplot() function results in the following plot?

Changing 'beside=TRUE' changes each column to be next to each other rather than stacked. 

>Q5. Generating all pairwise plots may help somewhat. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

The following creates a matrix of scatter plots to help understand the pairwise relationship between the different variables in the data set. If a given point lies on the diagonal for a given plot, it means that the same amount of people eat the same food in the different countries. 

```{r}
pairs(x, col= rainbow(10), pch= 16)
```
>Q6. What is the main difference between N. Ireland and the other countries of the UK in terms of this data set?

N. Ireland generally consumes less fresh fruit and eats more potatoes and soft drinks than the other countries of the UK. 

## PCA to the rescue
The main base R PCA function is called 'prcomp()' and we will need to give it the transpose of our input data!

```{r}
pca <- prcomp( t(x))
summary(pca)
```

```{r}
attributes(pca)
```

To make our new PCA plot (a.k.a. PCA score plot) we access 'pca$x'.

>Q7. Complete the code below to get the plot of PC1 vs PC2. The second line adds text labels over the data points. 

```{r}
plot(pca$x[,1], pca$x[,2])
text(pca$x[,1], pca$x[,2], colnames(x))
```

>Q8. Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at the start of this document. 

```{r}
country_cols <-c("orange", "red", "blue", "green")
plot(pca$x[,1], pca$x[,2], xlab= "PC1", ylab= "PC2")
text(pca$x[,1], pca$x[,2], colnames(x), col= country_cols)
```

Use square of pca$sdev to calculate how much variation in the original data each PC accounts to 

```{r}
v <- round( pca$sdev^2/sum(pca$sdev^2) * 100 )
v
```

Summarize in a plot of variances

```{r}
z <- summary(pca)
z$importance
```

```{r}
barplot(v, xlab="Principal Component", ylab="Percent Variation")
```

## Digging deeper (variable loadings)

Focus on PC1 as it accounts for >90% of variance

```{r}
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,1], las=2 )
```

>Q9. Generate a similar 'laodings plot' for PC2. What two food groups feature prominantely and what does PC2 mainly tell us about? 

```{r}
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,2], las=2 )
```

PC2 mainly shows how N. Ireland eats less fresh fruits and more potatoes and soft drinks. 

## Biplots

```{r}
biplot(pca)
```

## PCR of RNA-seq data

```{r}
url2 <- "https://tinyurl.com/expression-CSV"
rna.data <- read.csv(url2, row.names=1)
head(rna.data)
```
>Q10. How many genes and samples are in this data set?

```{r}
dim(rna.data)
```

There are 100 genes and 10 samples in this data set.

Take the transpose of our data 
```{r}
pca <- prcomp(t(rna.data), scale=TRUE)
```

This is a summary of how well PCA is doing
```{r}
summary(pca)
```
Do our PCA plot of this RNA-seq data 
```{r}
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2")
text(pca$x[,1], pca$x[,2], colnames(rna.data))
```

Quick barplot summmary of Proportion of Variance for each PC 
```{r}
plot(pca, main="Quick scree plot")
```

Variance captured per PC 
```{r}
pca.var <- pca$sdev^2
```

Percent variance 
```{r}
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)
pca.var.per
```

Generate scree-plot
```{r}
barplot(pca.var.per, main="Scree Plot", 
        names.arg = paste0("PC", 1:10),
        xlab="Principal Component", ylab="Percent Variation")
```

Make main PCR plot more attractive and useful 
```{r}
colvec <- colnames(rna.data)
colvec[grep("wt", colvec)] <- "red"
colvec[grep("ko", colvec)] <- "blue"

plot(pca$x[,1], pca$x[,2], col=colvec, pch=16,
     xlab=paste0("PC1 (", pca.var.per[1], "%)"),
     ylab=paste0("PC2 (", pca.var.per[2], "%)"))

text(pca$x[,1], pca$x[,2], labels = colnames(rna.data), pos=c(rep(4,5), rep(2,5)))
```

## Using ggplot

```{r}
library(ggplot2)

df <- as.data.frame(pca$x)

ggplot(df) + 
  aes(PC1, PC2) + 
  geom_point()
```

Add specific color and sample label aesthetics for wild-type and knock-out samples
```{r}
df$samples <- colnames(rna.data) 
df$condition <- substr(colnames(rna.data),1,2)

p <- ggplot(df) + 
        aes(PC1, PC2, label=samples, col=condition) + 
        geom_label(show.legend = FALSE)
p
```

Add some spit and polish 
```{r}
p + labs(title="PCA of RNASeq Data",
       subtitle = "PC1 clealy seperates wild-type from knock-out samples",
       x=paste0("PC1 (", pca.var.per[1], "%)"),
       y=paste0("PC2 (", pca.var.per[2], "%)"),
       caption="BIMM143 example data") +
     theme_bw()
```

